Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2020-05-04 manuscript/manuscript.tex
updated ms

  \caption{Forward-in-time simulations of demonstrate how temporal covariance,
    $G(t)$ trajectories, and convergence correlations arise during optima
    shifts of two different magnitudes, under Gaussian stabilizing selection.
    (A) Trait means across 30 replicate before and after optima shifts (solid
    lines), for two different magnitudes (indicated by color). The optimal
    trait values are indicated by the purple and yellow dashed lines.  (B) Mean
    temporal covariance $\cov(\Delta p_5, \Delta p_t)$ across 30 simulation
    replicates, where $t$ varies along the x-axis (points), with a
    loess-smoothed average (solid line). (C) $G(t)$ trajectories through time,
    for 30 replicate simulations across two optima shifts. The solid line is a
    loess-smoothed average. (D) The convergence correlations between two
    populations split from a common population, that underwent either an optima
    shift in the same direction (converge) and opposite directions (diverge)
    at generation five. In subfigures (B), (C), and (D), directional selection
    begins at generation five, when the optima shifts; this is indicated by the
    vertical dashed red line.}

@@

To remove systematic covariances in allele frequency change caused by tracking
the reference or minor allele, we randomly choose which allele's frequency to
track for each locus. Then, we calculate the variance-covariance matrix of
allele frequency changes using a Python software package we have written,
available at \url{http://github.com/vsbuffalo/cvtk}. This simultaneously
calculates temporal variances and covariances, and replicate covariances and
uses the sampling depth and number of diploid individuals to correct for bias
in the variance estimates and a bias that occurs in covariance estimates
between adjacent timepoints due to shared sampling noise (see Supplementary
Material Sections \ref{supp:ind-depth-var-corr}, \ref{supp:cov-corr}, and
\ref{supp:matrix-correction} for mathematical details of these estimators). We
assess that our bias correction procedure is working adequately through a
series of diagnostic plots that ensure that the procedure removes the
relationship between sampling depth and uncorrected variance and covariances
(Supplementary Figure \ref{suppfig:bergland-correction}).

@@

\subsection{Forward-in-time Simulations}

>>>>

  \caption{Forward-in-time simulations of demonstrate how temporal covariance,
    $G(t)$ trajectories, and convergence correlations arise during optima
    shifts of two different magnitudes, under Gaussian stabilizing selection.
    (A) Trait means across 30 replicate before and after optima shifts (solid
    lines), for two different magnitudes (indicated by color). The optimal
    trait values are indicated by the purple and yellow dashed lines.  (B) Mean
    temporal covariance $\cov(\Delta p_5, \Delta p_t)$ across 30 simulation
    replicates, where $t$ varies along the x-axis (points), with a
    loess-smoothed average (solid line). (C) $G(t)$ trajectories through time,
    for 30 replicate simulations across two optima shifts. The solid line is a
    loess-smoothed average. (D) The convergence correlations between two
    populations (each $1000$ diploids) split from a common population, that
    underwent either an optima shift in the same direction (converge) and
    opposite directions (diverge) at generation five. In subfigures (B), (C),
    and (D), directional selection begins at generation five, when the optima
    shifts; this is indicated by the vertical dashed red line.}

@@

To remove systematic covariances in allele frequency change caused by tracking
the reference or minor allele, we randomly choose which allele's frequency to
track for each locus. Then, we calculate the variance-covariance matrix of
allele frequency changes using a Python software package we have written,
available at \url{http://github.com/vsbuffalo/cvtk}. This simultaneously
calculates temporal variances and covariances, and replicate covariances and
uses the sampling depth and number of diploid individuals to correct for bias
in the variance estimates and a bias that occurs in covariance estimates
between adjacent timepoints due to shared sampling noise (see Supplementary
Material Sections \ref{supp:ind-depth-var-corr}, \ref{supp:cov-corr}, and
\ref{supp:matrix-correction} for mathematical details of these estimators). We
assess that our bias correction procedure is working adequately through a
series of diagnostic plots that ensure that the procedure removes the
relationship between sampling depth and uncorrected variance and covariances
(Supplementary Figure \ref{suppfig:bergland-correction}).  \vb{Through our
  simulations we find that our estimates can differ based on how fixations and
  losses are handled in long time-series (Supplementary Material Section
  \ref{supp:fixation}) but none of our findings in the main text are altered by
  this decision (Supplementary Material Figures \ref{suppfig:supp-fig-1-nofix}
  and \ref{suppfig:supp-fig-3-nofix}).}

@@

\subsection{Forward-in-time Simulations}
\label{supp:forward}
\vb{
To explore how aspects of genetic architecture, model of selection, and
experimental design impact temporal covariance, the $G(t)$ trajectories, and
convergence correlations, we ran extensive forward-in-time using SLiM
\parencite{Haller2019-vu}; here we discuss the Gaussian Stabilizing Selection
simulations in Figured \ref{fig:figure-4}, but Supplementary Materials Section
\ref{supp:forward} describes these simulation routines in much more detail as
well as additional simulations we have conducted.

We simulated directional selection on a trait by first burning in a population
of $N = 1000$ diploids under GSS for $10N$ generations with the stabilizing
selection trait variance $V_s = 1$ and an optima set at zero. We targeted a
polygenic architecture by setting the trait mutation rate to $10^{-8}$ per
basepair, per generation, in addition to having a separate neutral mutation of
$10^{-8}$ which created neutral mutations which we used to calculate the
temporal covariances. Our simulated region was 50 megabases in length (about
one quarter of a \emph{Drosophila} chromosome), and trait alleles were randomly
selected to have a $\pm 0.01$ effect size. By tracking the trait mean through
the burnin, we found it converged as expected. After a $10N$ burnin, the
population was split into two different replicate population sizes, to capture
the effect of bottlenecks in selection experiments (these population sizes
were, 50, 500, and 1000 diploids; the later representing no bottleneck). Each
population then underwent an optima shift of either 0.1, 0.5, or 1 on
generation five, with the first four generations serving as a control. These
optima shifts were either in the same direction (converging), different
directions (diverging), or only one optima shifted (as a control). By tracking
the trait mean, we saw that it converged as expected during burnin, and had the
appropriate response to selection (Supplementary Material Figure
\ref{suppfig:gss-zbar}). Using the neutral population frequency data from these
simulations, we calculated the temporal covariances, $G(t)$ trajectories, and
convergence correlations.
}
